Notes

### 2022-09-05 chapters/introduction.tex
Finish document

Under a functional framework, the objects of interest are \textit{random functions} instead of random points in a finite-dimensional space. While in principle the functional data could be simply regarded as a discretized vector in a very high dimension (an indeed such a discretization is performed in practice), there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. Thus, the general idea is to assume the existence of an underlying sufficiently smooth function that generates each (possibly noisy) functional observation, even though we only record it on a finite grid of points.

@@

\begin{center}
\color{teal}\FourStar
\end{center}

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. Moreover, we will concentrate our efforts on the case in which the response is a scalar, though function-on-function and function-on-scalar regression are also interesting scenarios widely explored in the literature. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\) (i.e. a process with finite second moments and whose realizations are square-integrable functions indexed on \([0,1]\)). The underlying probability space \((\Omega, \mathcal A, \P)\) is not important. We will further suppose that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\D_n =\{(x_i, y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), where the functional observations are recorded on a common finite grid \(\{t_j\}\subset [0, 1]\). Our ultimate aim will be to accurately predict the response corresponding to unlabeled samples from \(X\). Figure~\ref{fig:linear_data_example} depicts a typical data set used in functional regression, and we already saw in Figure~\ref{fig:tecator} what a functional classification data set may look like.

>>>>

Under a functional framework, the objects of interest are \textit{random functions} instead of random points in a finite-dimensional space. While in principle the functional data could be simply regarded as a discretized vector in a very high dimension (and indeed such a discretization is performed in practice), there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. Thus, the general idea is to assume the existence of an underlying sufficiently smooth function that corresponds to each (possibly noisy) functional observation, even though we only record it on a finite grid of points.

@@

\begin{center}
  \vspace{.25em}
\color{teal}\FourStar
\end{center}

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. Moreover, we will concentrate our efforts on the case in which the response is a scalar, though function-on-function and function-on-scalar regression are also interesting scenarios widely explored in the literature. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\) (i.e. a process with finite second moments and whose realizations are square-integrable functions indexed on \([0,1]\)). The underlying probability space \((\Omega, \mathcal A, \P)\) is not important. We will further suppose that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), where the functional observations are recorded on a common finite grid \(\{t_j\}\subset [0, 1]\). Our ultimate aim will be to accurately predict the response corresponding to unlabeled samples from \(X\). Figure~\ref{fig:linear_data_example} depicts a typical data set used in functional regression, and we already saw in Figure~\ref{fig:tecator} what a functional classification data set may look like.

### 2022-09-05 chapters/model-choice.tex
Finish document

In this chapter we gather together several remarks on the main choices made throughout the design of our model, as well as some examples of validation strategies that attempt to measure the goodness-of-fit of the model given the observed data.

@@

The most straightforward solutions consist on imposing an artificial identifiability constraint on the parameters to break the symmetry of their posterior distributions; see \citet{jasra2005markov} and references therein. A common approach that seems to work well is to simply impose an ordering in the offending parameters, which in our case would mean requiring for example that \(\beta_i < \beta_j\) for \(i < j\), or the analogous with the \(\tau_j\). We have implemented a variation of this method described in \citet{simola2021approximate}, which works by post-processing the samples and relabeling the components to satisfy the order constraint mentioned above, choosing either \(b\) or \(\tau\) depending on which set of ordered parameters would produce the largest separation between any two of them (suitably averaged across all iterations of the chains).

@@

One of the key decisions in our Bayesian modeling scheme was whether to consider the number of components \(p\) as a member of the parameter space and integrate it into the model. While theoretically we can impose a prior distribution on \(p\) as well (e.g. a categorical distribution with a fixed maximum value), we found that it would have some unwanted practical implications. For instance, it would make the implementation more complex, since the dimensionality of the parameters \(b\) and \(\tau\) would need to be fixed at a certain maximum value beforehand, but the working value of \(p\) within the MCMC algorithm would vary from one iteration to the next. In this case we would have no immediate way of tracking down which set of parameters is ``active'' at any given time. A simple approach would be to always consider the first \(p\) parameters and ignore the rest, and we did indeed try this technique, but it gave rise to new difficulties and the results obtained were not good. In fact, the label switching issue is accentuated when \(p\) is allowed to vary \citep[c.f.][Sec.~2.3]{grollemund2019bayesian}, and on top of that, the interpretation of, say, the first coefficient \(\beta_1\) in a model with \(3\) components is different than the interpretation of the same coefficient in a model with only \(2\) components.

@@

As for the default values of the rest of hyperparameters in the prior distributions in~\eqref{eq:prior-linear}, several comments are in order:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use the MLE of \(b\). Although the likelihood function is rather involved, an approximation of the optimal value is enough for our purposes. Our numerical studies suggest that the results are much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation indicates that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\), where \(\lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\) is the largest eigenvalue of the matrix \(\mathcal X_\tau'\mathcal X_\tau\), and \(\tilde\eta > 0\) is the actual tuning parameter (which can be selected for example by cross-validation strategies). This standardization technique has been used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}

@@

The MCMC method chosen for approximating the posterior distribution in our models is the affine-invariant ensemble sampler described in Section~\ref{sec:mcmc}. As mentioned there, we use the implementation in the Python library \textit{emcee}, which is both reliable and easy to use; it aims to be a general-purpose package that performs well in a wide class of problems. One advantage of this method, apart from the property of affine-invariance, is that it only requires us to specify a few hyperparameters, irrespective of the underlying dimension. This contrasts to, say, the \(\mathcal O(N^2)\) degrees of freedom corresponding to the covariance matrix of an \(N\)-dimensional jump distribution in Metropolis-Hastings.
Furthermore, following \citet{goodman2010ensemble} we can use the \textit{integrated autocorrelation time} to compute an estimate of the effective sample size (i.e. the number of independent samples) after the sampling is done.

After selecting the number of iterations and the number of chains we want, we need to specify the initial points for each of them. As pointed out in \citet{foreman2013emcee}, a good initial choice is a Gaussian ball around a point in \(\Theta_p\) that is expected to have a high probability with respect to the objective distribution. In our implementation we adopt this method, choosing an approximation of the MLE of \(\theta\) as the central point in each case. To perform this approximation we employ the optimization suite of the Python library \textit{scipy}\footnote{\url{https://docs.scipy.org/doc/scipy/}}, and in particular we use the Basin-hopping algorithm \citep{wales1997global}. This is a two-phase stochastic method that combines global steps with local optimization, in the hope of avoiding getting stuck too quickly in local maxima. To reduce the effects of randomness, we run the algorithm a few times and retain the point with the highest likelihood, and to avoid biasing the sampler too much towards the specific point selected, we let a fraction of the initial points be random (within reasonable bounds).

@@

\begin{figure}[ht!]
  \centering
    \setlength{\fboxrule}{.5pt}%
  \fbox{\includegraphics[width=.9\textwidth]{credible_intervals}}
  \caption{Example of 94\% credible intervals on the posterior distribution of the impact points}\label{fig:credible_intervals}
\end{figure}

>>>>

In this chapter we gather together several remarks on the main choices made during the design and implementation of our model, as well as some examples of validation strategies that attempt to measure the goodness-of-fit of the model given the observed data.

@@

The most straightforward solutions consist on imposing an artificial identifiability constraint on the parameters to break the symmetry of their posterior distributions; see \citet{jasra2005markov} and references therein. A common approach that seems to work well is to simply impose an ordering in the parameters in question, which in our case would mean requiring for example that \(\beta_i < \beta_j\) for \(i < j\), or the analogous with \(\tau\). We have implemented a variation of this method described in \citet{simola2021approximate}, which works by post-processing the samples and relabeling the components to satisfy the order constraint mentioned above, choosing either \(b\) or \(\tau\) depending on which set of ordered parameters would produce the largest separation between any two of them (suitably averaged across all iterations of the chains).

@@

One of the key decisions in our Bayesian modeling scheme was whether to consider the number of components \(p\) as a member of the parameter space and integrate it into the model. While theoretically we could impose a prior distribution on \(p\) as well (e.g. a categorical distribution with a fixed maximum value), we found that it would have some unwanted practical implications. For instance, it would make the implementation more complex, since the dimensionality of the parameters \(b\) and \(\tau\) would need to be fixed at a certain maximum value beforehand, but the working value of \(p\) within the MCMC algorithm would vary from one iteration to the next. In this case we would have no immediate way of tracking down which set of parameters is ``active'' at any given time. A simple approach would be to always consider the first \(p\) parameters and ignore the rest, and we did indeed try this technique, but it gave rise to new difficulties and the results obtained were not good. In fact, the label switching issue is accentuated when \(p\) is allowed to vary \citep[c.f.][Sec.~2.3]{grollemund2019bayesian}, and on top of that, the interpretation of, say, the first coefficient \(\beta_1\) in a model with \(3\) components is different than the interpretation of the same coefficient in a model with only \(2\) components.

@@

As for the default values of the rest of hyperparameters in the prior distributions in~\eqref{eq:prior-linear}, several comments are in order:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use the MLE of \(b\). Although the likelihood function is rather involved, an approximation of the optimal value is enough for our purposes. Our numerical studies suggest that the results are much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation indicates that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\), where \(\lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\) is the largest eigenvalue of the matrix \(\mathcal X_\tau'\mathcal X_\tau\), and \(\tilde\eta > 0\) is the actual tuning parameter (which can be selected for instance by cross-validation strategies). This standardization technique has been used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}

@@

The MCMC method chosen for approximating the posterior distribution in our models is the affine-invariant ensemble sampler described in Section~\ref{sec:mcmc}. As mentioned there, we utilize the computational implementation in the Python library \textit{emcee}, which is both reliable and easy to use; it aims to be a general-purpose package that performs well in a wide class of problems. One advantage of this method, apart from the property of affine-invariance, is that it only requires us to specify a few hyperparameters, irrespective of the underlying dimension. This contrasts to, say, the \(\mathcal O(N^2)\) degrees of freedom corresponding to the covariance matrix of an \(N\)-dimensional jump distribution in Metropolis-Hastings.
Furthermore, following \citet{goodman2010ensemble} we can use the \textit{integrated autocorrelation time} to compute an estimate of the effective sample size (i.e. the number of independent samples) after the sampling is done.

After selecting the number of iterations and the number of chains we want, we need to specify the initial points for each of them. As pointed out in \citet{foreman2013emcee}, a good initial choice is a Gaussian ball around a point in \(\Theta_p\) that is expected to have a high probability with respect to the objective distribution. In our implementation we adopt this method, choosing an approximation of the MLE of \(\theta\) as the central point in each case. To perform this approximation we employ the optimization suite of the Python library \textit{scipy}\footnote{\url{https://docs.scipy.org/doc/scipy/}}, and in particular we use the Basin-hopping algorithm \citep{wales1997global}. This is a two-phase stochastic method that combines global steps with local optimization, in the hope of avoiding getting stuck too quickly in local maxima. To reduce the effects of randomness, we run the algorithm a few times and retain the point with the highest likelihood, and to avoid biasing the sampler too much towards the specific point selected, we let a fraction of the initial points be random (within reasonable bounds). This approximation is also used to specify the hyperparameter \(b_0\).

@@

\begin{figure}[ht!]
  \centering
    \setlength{\fboxrule}{.5pt}%
  \fbox{\includegraphics[width=.9\textwidth]{credible_intervals}}
  \caption{Example of 94\% credible intervals on the posterior distribution of the impact points.}\label{fig:credible_intervals}
\end{figure}
